### 2024-06-11 paper/tex/01-introduction.tex
corrections

The most famous of the algorithms for handling this is the Damas-Hindley-Milner
(DHM) algorithm \cite{hindley69:principal,milner78:theory,damas82:principal}.
DHM deals with this choice using \emph{unification}. Whenever we would have had
to introduce a particular concrete type in the specification, the DHM algorithm
introduces a \emph{unification variable}, and incrementally instantiates this variable
as more and more type constraints are observed.

However, the universal quantifier is not the only quantifier! Dual to the
universal quantifier $\forall$ is the existential quantifier $\exists$. Even
though existential types have an equal logical status to universal types, they
have been studied much less frequently in the literature. The most widely-used
algorithm for handling existentials is the algorithm of Odersky and Laufer
\cite{laufer94:polymorphic}. This algorithm treats existentials in a
second-class fashion: they are not explicit connectives of the logic but rather
are tied to datatype declarations. As a result, packing and unpacking
existential types is tied to constructor introductions and constructor
eliminations in pattern matches. This allows Damas-Milner inference to continue
to work almost unchanged, but it does come at the cost of losing first-class
existentials and also of restricting the witness types to monomorphic types.

There has been a limited amount of work on support for existential types as a
first-class concept. In an unpublished draft, \citet{leijen06:first-class}
studied an extension of Damas-Milner inference in which type schemes contain
alternating universal and existential quantifiers. Quantifiers still range over
monotypes (\ie types with no quantifiers inside of them), and higher-rank 
polymorphism is not permitted: one cannot instantiate a quantifier with a type
that itself contains quantifiers.
More recently,
\citet{dunfield16:existential} studied type inference for existential types in
the context of GADT type inference, which, while still predicative, supported
higher-rank types (i.e., quantifiers can occur anywhere inside of a type
scheme). \citet{eisenberg21:existential} propose a system which only permits
types of the form forall-exists, but which permits projective elimination in the
style of ML modules.

All of these papers are restricted to \emph{predicative} quantification, where
quantifiers can only be instantiated with monotypes. However, existential types
in full System F are \emph{impredicative}---that is, quantifiers can be
instantiated with arbitrary types, specifically including types containing
quantifiers.

Historically, inference for impredicative quantification has been neglected, due
to results by \citet{tiuryn-urzczyn-96} and \citet{chrzaszcz-98} which show that
not only is full type inference for \systemf undecidable, but even that the
subtyping relation induced by type instantiation is undecidable. However, in
recent years, interest in impredicative inference has revived (for example, the
work of \citet{serrano-2020}), with a focus on avoiding the undecidability
barriers by doing \emph{partial} type inference. That is, we no longer try to
infer the type of any term in the language, but rather accept that the
programmer will need to write annotations in cases where inference would be too
difficult. Unfortunately, it is often difficult to give a specification for when
the partial algorithm succeeds---for example, \citet{eisenberg21:existential}
observe that their algorithm lacks a declarative specification, and explain why
existential types make this particularly difficult to do.

@@

To research the type inference in impredicative polymorphic systems in the
presence of existential types, we make the following
\emph{\textbf{contributions}}.

\paragraph{Declarative Type System Specification (\cref{sec:overview,sec:declarative-system})} 
    We devise \fexists---a second-order polymorphic type system that supports first-class
    existential and universal quantifiers, both of which can be instantiated
    impredicatively. To evade the undecidability results surrounding subtyping
    for \systemf, we base the system on \CBPV \cite{levy2006:cbpv}: we
    use \emph{polarization} of types and terms to formulate the restrictions
    that make typing decidable.

    We give a \emph{declarative} specification of local type inference in this
    system (\ie the typing information is only available at a limited scope)
    based on \emph{subtyping} relation. The specification makes it easy to see
    what is and what is not inferrable in the system.

@@

\paragraph{Type Inference Algorithm (\cref{sec:algorithm})} We give a type
  inference algorithm implementing the declarative specification. Because our
  system supports both existential and universal quantifiers, our implementation
  of the subtyping relation requires the use of \emph{anti-unification}---a dual
  to unification that finds the most specific generalization rather than the
  most general unifier. To our knowledge, anit-unification has never been used
  in the context of polymorphic type inference.

@@

\paragraph{Theoretical results: correctness of the algorithm (\cref{sec:proofs})}
  We prove that our algorithm is terminating and both sound and complete with
  respect to the declarative specification. This proof of soundness and completeness
  of the algorithm required us to introduce an intermediate system in-between
  the declarative and the algorithmic systems, and furthermore, the
  soundness and completeness proofs are mutually recursive with each other.
  The central theorems and the key proof ideas are formulated in \cref{sec:proofs}, 
  while the full definitions and formal proofs are deferred to the appendix (\cref{appendix}).

% \paragraph{Extrapolation of the inference framework to other type systems (\cref{sec:extensions})}
%     We explore the design space further to show how the
%     same type inference scheme can be applied to
%     work with different type systems. In particular,
%     \begin{enumerate*}
%       \item[(i)] in combination with \emph{bidirectional} typechecking---the 
%         approach used in the original local type inference paper of 
%         \citet{pierce2000:local} that minimizes the number of needed annotations;
%       \item[(ii)] in systems with different subtyping rules, such as  
%         the one used in \citet{zhao22:elementary}---a system that
%         permits explicit type applications;
%       \item[(iii)] in systems with \emph{bounded} polymorphic quantifiers.
%         That is, one can only instantiate a quantifier with a 
%         type that is a subtype of a given bound.
%     \end{enumerate*}

% \ilyam{Is the following needed since we have the contributions? I just wanted to introduce the appendix somewhere...}
% \paragraph{\textbf{Structure of this Paper}}
% In \cref{sec:overview}, we discuss the language of the \fexists system, providing
% examples, outlining limitations, and summarizing the main ideas behind the
% algorithm. \Cref{sec:declarative-system} presents a formal definition of the
% declarative specification and its key properties. In \cref{sec:algorithm}, the
% type inference algorithm is introduced, detailing a hierarchy of subroutines;
% upon perusing this section, one could \emph{implement} the type inference
% algorithm. \Cref{sec:proofs} overviews the algorithm's correctness proof,
% explaining its structure and how the declarative system's features make it
% decidable. \Cref{sec:extensions} covers potential \fexists extensions and future
% work directions. Finally, \cref{appendix} contains comprehensive formal
% definitions and proofs of lemmas and theorems used in the paper; in instances
% where the full definitions might not be of immediate interest, we direct the
% reader to the appendix.

% \paragraph{\textbf{Acknowledgments}} We would like to thank Meven
% Lennon-Bertrand for his extremely helpful comments and suggestions.

>>>>

The most famous of the algorithms for handling this is the Damas-Hindley-Milner
(DHM) algorithm \cite{hindley69:principal,milner78:theory,damas82:principal}.
DHM deals with this choice using \emph{unification}. Whenever we would have had
to guess a particular concrete type in the specification, the DHM algorithm
introduces a \emph{unification variable}, and incrementally instantiates this variable
as more and more type constraints are observed.

However, the universal quantifier is not the only quantifier! Dual to the
universal quantifier $\forall$ is the existential quantifier $\exists$. Even
though existential types have an equal logical status to universal types, they
have been studied much less frequently in the literature. The most widely-used
algorithm for handling existentials is the algorithm of Odersky and L{\"a}ufer
\cite{laufer94:polymorphic}. This algorithm treats existentials in a
second-class fashion: they are not explicit connectives of the logic but rather
are tied to datatype declarations. As a result, packing and unpacking
existential types is tied to constructor introductions and constructor
eliminations in pattern matches. This allows Damas-Milner inference to continue
to work almost unchanged, but it does come at the cost of losing first-class
existentials and also of restricting the witness types to monomorphic types
(\ie types with no quantifiers inside of them).

There has been a limited amount of work on support for existential types as a
first-class concept. In an unpublished draft, \citet{leijen06:first-class}
studied an extension of Damas-Milner inference in which type schemes contain
alternating universal and existential quantifiers. Quantifiers still range over
monotypes, so higher-rank 
polymorphism is not permitted: one cannot instantiate a quantifier with a type
that itself contains quantifiers.
More recently,
\citet{dunfield16:existential} studied type inference for existential types in
the context of GADT type inference, which, while still predicative, supported
higher-rank types (i.e., quantifiers can occur anywhere inside of a type
scheme). \citet{eisenberg21:existential} propose a system which only permits
types of the form forall-exists, but which permits projective elimination in the
style of ML modules.

All of these papers are restricted to \emph{predicative} quantification, where
quantifiers can only be instantiated with monotypes. However, existential types
in full \systemf are \emph{impredicative}---that is, quantifiers can be
instantiated with arbitrary types, specifically including types containing
quantifiers.

Historically, inference for impredicative quantification has been neglected, due
to results by \citet{tiuryn-urzczyn-96} and \citet{chrzaszcz-98} which show that
not only is full type inference for \systemf undecidable, but even that the
subtyping relation induced by type instantiation is undecidable. However, in
recent years, interest in impredicative inference has revived (for example, the
work of \citet{serrano-2020}), with a focus on avoiding the undecidability
barriers by doing \emph{partial} type inference. That is, we no longer try to
infer the type of any term in the language, but rather accept that the
programmer will need to write annotations in cases where inference would be too
difficult. Unfortunately, it is often difficult to give a specification for when
the partial algorithm succeeds---for example, \citet{eisenberg21:existential}
observe that their algorithm lacks a declarative specification, and explain why
supporting existential types makes giving a specification particularly
difficult.

@@

We make the following \emph{\textbf{contributions}}:

\paragraph*{Declarative Type System Specification (\cref{sec:overview,sec:declarative-system})} 
    We present \fexists---a second-order polymorphic type system that supports first-class
    existential and universal quantifiers, both of which can be instantiated
    impredicatively. To evade the undecidability results surrounding subtyping
    for \systemf, we base the system on \CBPV \cite{levy2006:cbpv}: we
    use \emph{polarization} of types and terms to formulate the restrictions
    that make typing decidable.

    We give a \emph{declarative} specification of local type inference in this
    system (\ie the typing information is only available at a limited scope)
    based on a \emph{subtyping} relation. Our specification makes it easy to see
    what is and what is not inferrable in the system.

@@

\paragraph*{Type Inference Algorithm (\cref{sec:algorithm})} We give a type
  inference algorithm implementing the declarative specification. Because our
  system supports both existential and universal quantifiers, our implementation
  of the subtyping relation requires the use of \emph{anti-unification}---a dual
  to unification that finds the most specific generalization rather than the
  most general unifier. To our knowledge, anti-unification has never been used
  in the context of polymorphic type inference.

@@

\paragraph*{Correctness of the algorithm (\cref{sec:proofs})}
  We prove that our algorithm is terminating and both sound and complete with
  respect to the declarative specification. This proof of soundness and completeness
  of the algorithm required us to introduce an intermediate system in-between
  the declarative and the algorithmic systems, and furthermore, the
  soundness and completeness proofs are mutually recursive with each other.
  The central theorems and the key proof ideas are formulated in \cref{sec:proofs}, 
  while the full definitions and formal proofs are deferred to the appendix (\cref{appendix}).

% \paragraph*{Extrapolation of the inference framework to other type systems (\cref{sec:extensions})}
%     We explore the design space further to show how the
%     same type inference scheme can be applied to
%     work with different type systems. In particular,
%     \begin{enumerate*}
%       \item[(i)] in combination with \emph{bidirectional} typechecking---the 
%         approach used in the original local type inference paper of 
%         \citet{pierce2000:local} that minimizes the number of needed annotations;
%       \item[(ii)] in systems with different subtyping rules, such as  
%         the one used in \citet{zhao22:elementary}---a system that
%         permits explicit type applications;
%       \item[(iii)] in systems with \emph{bounded} polymorphic quantifiers.
%         That is, one can only instantiate a quantifier with a 
%         type that is a subtype of a given bound.
%     \end{enumerate*}

% \ilyam{Is the following needed since we have the contributions? I just wanted to introduce the appendix somewhere...}
% \paragraph*{\textbf{Structure of this Paper}}
% In \cref{sec:overview}, we discuss the language of the \fexists system, providing
% examples, outlining limitations, and summarizing the main ideas behind the
% algorithm. \Cref{sec:declarative-system} presents a formal definition of the
% declarative specification and its key properties. In \cref{sec:algorithm}, the
% type inference algorithm is introduced, detailing a hierarchy of subroutines;
% upon perusing this section, one could \emph{implement} the type inference
% algorithm. \Cref{sec:proofs} overviews the algorithm's correctness proof,
% explaining its structure and how the declarative system's features make it
% decidable. \Cref{sec:extensions} covers potential \fexists extensions and future
% work directions. Finally, \cref{appendix} contains comprehensive formal
% definitions and proofs of lemmas and theorems used in the paper; in instances
% where the full definitions might not be of immediate interest, we direct the
% reader to the appendix.

\paragraph*{\textbf{Acknowledgments}} We would like to thank Meven
Lennon-Bertrand for his extremely helpful comments and suggestions.

### 2024-06-11 paper/tex/02-overview.tex
corrections

In the presence of impredicative polymorphism, the standard \systemf has
undecidable subtyping and type inference. It means that any inference algorithm
is incomplete: it only infers the types for a \emph{fragment} of \systemf.

@@

Every term and type fo \fexists has either \emph{positive} or \emph{negative}
polarity. The type variables are annotated with their polarities (e.g. $[[α⁺]]$
or $[[β⁻]]$), and the types can change polarity via the shift operators $[[↑]]$
(positive to negative) and $[[↓]]$ (negative to positive).

Positive expressions correspond to \emph{values} in the \CBPV system, and
negative expressions correspond to \emph{computations}. The difference between
the computations and values is intuitively described in the motto of \CBPV: ``a
value is, a computation does''. In particular, an argument of a function is
always a value (\ie positive), and the resulting type, as well as the function
itself, is a computation (\ie negative). Analogously, the polymorphic
$\forall$ quantifies a computation over positive types.

@@

Generally, any term and type of \systemf can be embedded into \fexists in
multiple ways. This is because at any given point, one can opt for either the
positive or negative polarity of the subterm. We will delve into the consistent
term-level and type-level embeddings (\ie preserving typing) in more detail in
\cref{sec:rel-to-systemf}. \Cref{fig:polarization-examples} showcases the
polarized versions of the standard \systemf terms. Next, we will discuss several
examples (using the terms from \cref{fig:polarization-examples}) to illustrate
what can and cannot be inferred in \fexists.

\begin{figure}[h]
  \begin{align*}
    [[map]] &: [[↓(∀α⁺.∀β⁺.↓(α⁺ → ↑β⁺) → List α⁺ → ↑List β⁺)]] \\
    [[len]] &: [[↓(∀α⁺.List α⁺ → ↑Int)]] \\
    [[choose]] &: [[↓(∀α⁺.α⁺ → α⁺ → ↑α⁺)]] \\
    [[id]] &: [[↓(∀α⁺.α⁺ → ↑α⁺)]] \\
    [[auto]] &: [[↓(↓(∀α⁺.α⁺ → ↑α⁺) → (∀α⁺.α⁺ → ↑α⁺))]]
  \end{align*}
  \caption{Polarization of standard \systemf terms}
  \label{fig:polarization-examples}
\end{figure}

\paragraph{Application of a Polymorphic Function}
  A polymorphic function can be applied to a value of a concrete type. In this
  case, the quantified variables are instantiated to make the application
  type-check. For instance, the polymorphic $[[len]]$ can be applied to a list
  of any positive type, and the resulting type will be inferred by the
  algorithm:
  $$[[· ; Φ ⊢ let x = len([one,two,three]); return x : ↑Int]]$$

\paragraph{Polymorphic Self-application}
  Unlike predicative systems, \fexists permits self-application of polymorphic
  functions. For instance, $[[id]]$ can be applied to itself resulting in a
  suspended polymorphic computation of type $[[↓(∀α⁺.α⁺ → ↑α⁺)]]$, which, in
  turn, can be applied to itself:
  $$[[· ; Φ ⊢ let x = id(id); let y = x(x); return y : ↑↓(∀α⁺.α⁺ → ↑α⁺)]]$$

\paragraph{Non-example: Polymorphic Arguments}
The focus of our system is the \emph{local} type inference
\cite{pierce2000:local}, which has certain limitations. In particular, the
instantiation does not happen when the polymorphic term is in the argument
position. For example, one could expect the following to be inferrable:
\mbox{$[[· ; Φ ⊢ let x = map(id, [two, three, nine]); return x : ↑ List Int]]$}.
However, to infer this, $[[id]]$ must be instantiated to \mbox{$[[Int → ↑Int]]$}
which requires the information to be propagated from the neighboring branch of
the syntax tree (\mbox{$[[ [two, three, nine] ]]$}), \ie bypass the locality.

The annotated version of the same term infers the type successfully, 
since all the polymorphic variables of the arguments are instantiated:
$$[[· ; Φ ⊢ let x = map((id : ↓(Int → ↑Int)), [two, three, nine]); return x : ↑ List Int]]$$

\paragraph{Inferring the Common Supertype}
  The function $[[choose]]$ takes two arguments of the same
  polymorphic type. To apply it to two values of different (concrete) types, 
  the inference algorithm must find the minimal type they can be coerced 
  to---the least common supertype.

@@

\paragraph{Inferring a general type}
  As mentioned before, the local inference does not instantiate the polymorphic
  types in the argument position.
  Because of that, the application 
  \mbox{$[[let x = choose (id, auto); return x ]]$} cannot infer 
  $[[↑↓(↓(∀α⁺.α⁺ → ↑α⁺) → (∀α⁺.α⁺ → ↑α⁺))]]$.
  However, the inference succeeds and infers a less informative type:
  $$[[· ; Φ ⊢ let x = choose (id, auto); return x : ↑ ∃γ⁻.↓γ⁻]]$$

@@

\paragraph{Syntactic Conventions}
We always consider terms and types up to alpha-equivalence, and capture-avoiding
substitution. Besides, we assume the quantification is associative,
for example, $[[∀pas.∀pbs.iN]]$ and $[[∀pas,pbs.iN]]$
represent the same type.

\paragraph{Type Context and Type Well-formedness}

@@

\paragraph{Missing constructors}
Notice that the language does not have first-class applications: 
their role is played by the applicative let forms, binding 
the result of a \emph{fully applied} function to a variable.
Also notice that the language does not have a type application (i.e. the eliminator of $[[∀]]$) and dually, it does not have \pack (i.e. the constructor of $[[∃]]$).
This is because the instantiation of polymorphic and existential types is inferred by the algorithm. 
In \cref{sec:extensions}, we discuss the way to modify the system to introduce \emph{explicit} type applications.

@@

The inference algorithm for \fexists is \emph{local}: it has a limited scope of
the inference information and does not propagate it between far-apart branches
of the syntax tree. Nevertheless, many difficulties appear in the inference of
the combination of polymorphic and existential types. Next, we discuss the most
challenging of these difficulties and how they are addressed in the algorithm.

\paragraph{Subtyping} 
The inference algorithm's complexity mainly lies in the subtyping relation,
which is, in particular, responsible for polymorphic instantiation. It's this
subtyping relation that enables us to apply polymorphic terms in situations
where their specific instantiations are anticipated.

@@

\paragraph{Equivalence}
The system introduces another complexity: the equivalence induced by subtyping
($[[Γ ⊢ iN ≈ iM]] \defiff [[Γ ⊢ iN ≤ iM]] \text{ and } [[Γ ⊢ iM ≤ iN]]$) is not
straightforward. This relation extends beyond just alpha-equivalent types. As
the neighboring polymorphic quantifiers can be instantiated in arbitrary order,
the equivalence relation must permit the \emph{permutations} of the quantifiers.
For example, we have $[[· ⊢ ∀α⁺.∀β⁺.iN ≈ ∀β⁺.∀α⁺.iN]]$. Similarly, the
equivalence permits the \emph{removal/addition} of the unused quantifiers: $[[·
⊢ ∀α⁺.↑Int ≈ ↑Int]]$.

@@

\paragraph{Least Upper Bound}
Eventually, the subtyping algorithm reduces the initial subtyping problem to a set of
constraints that need to be resolved. At this moment, another problem arises:
as one variable can occur in multiple constraints, the resolution becomes non-trivial.
In particular, to resolve the conjunction of $[[α̂⁺ ≥ iP]]$ and $[[α̂⁺ ≥ iQ]]$,
we would need to find the least upper bound of $[[iP]]$ and $[[iQ]]$.

@@

To find this least upper bound, we use \emph{anti-unification}---the dual of
unification. While unification finds \emph{the most general instance} of two
given patterns, the anti-unification finds \emph{the most detailed} (under the
restrictions of the system) \emph{pattern} that matches two given types. In the
example above, the anti-unifier of $[[↓↑Int]]$ and $[[↓↑Bool]]$ is $[[↓α̂⁻]]$,
and in \fexists, this unifier is the most specific (in particular because
$[[↓↑α̂⁺]]$ is not considered as only \emph{negative} variables are
existentially quantified).

\paragraph{Impredicativity}
Another difficulty that hinders the constraint resolution arises from the
impredicativity of the system---the ability to quantify over the types that are 
themselves polymorphic. In the impredicative polymorphic system (even
with only $[[∀]]$-quantifiers), the naturally formulated subtyping is
undecidable, which follows from the undecidability of second-order unification.
All the more the undecidability is evident in the presence of existential types
and subtyping constraints. For instance, in \fexists the constraint 
$[[α̂⁺ :≥ ↓↑α̂⁺]]$ would have a surprising impredicative solution $[[α̂⁺]] = [[∃α⁻.↓α⁻]]$.

To maintain the decidability in the presence of impredicativity, we carefully
restrict the system using the polarity stratification. One of the main
constraints that we put on subtyping is the invariance of the shift operators:
the subtyping $[[↑iP ≤ ↑iQ]]$ is only allowed if $[[iQ ≥ iP]]$
\emph{and} $[[iP ≥ iQ]]$. These constraints do not trivialize the system: we
still have non-trivial upper bounds, and thus, need to use anti-unification to
find them. However, it prevents such examples as $[[α̂⁺ :≥ ↓↑α̂⁺]]$, and 
as we prove, makes the constraint resolution decidable.

>>>>

In the presence of impredicative polymorphism, \systemf has
undecidable subtyping and type inference. It means that any inference algorithm
is incomplete: it only infers the types for a \emph{fragment} of \systemf.

@@

Every term and type of \fexists has either \emph{positive} or \emph{negative}
polarity. The type variables are annotated with their polarities (e.g. $[[α⁺]]$
or $[[β⁻]]$), and the types can change polarity via the shift operators $[[↑]]$
(positive to negative) and $[[↓]]$ (negative to positive).

Positive expressions correspond to \emph{values} in the \CBPV system, and
negative expressions correspond to (potentially effectful) \emph{computations}.
The difference between the computations and values is intuitively described in
the motto of \CBPV: ``a value is, a computation does''. In particular, a
function type always takes a value and returns a computation. Similarly, the
polymorphic $\forall$ quantifies a computation over positive types.

Whenever a computation is used in a value position (e.g. as an argument to a
function), it must be \emph{suspended} by a thunk constructor, which at the
type level corresponds to the downshift operator $[[↓iN]]$.
Symmetrically, to use a value as a computation, one can \emph{return} it,
which at the type level is reflected as an up-shift $[[↑iP]]$.

@@

Generally, any term and type of \systemf can be embedded into \fexists in
multiple ways corresponding to different evaluation strategies of \systemf,
which we will cover in more detail in \cref{sec:rel-to-systemf}.

\Cref{fig:polarization-examples} illustrates how standard \systemf types are
polarized when call-by-value evaluation is presumed.

Observe the consistent polarization pattern of types. Consider the type for
$[[double]]$, which in \systemf would be $[[Int → Int]]$. In polarized form, it
becomes $[[↓(Int → ↑Int)]]$---a \emph{suspended} function (indicated by the
downshift) that accepts an integer $[[Int]]$ and yields an integer
\emph{computation} (indicated by the upshift) $[[↑Int]]$.

Moving on to the examples in \cref{fig:polarization-examples}, we will delve
into several cases that demonstrate the capabilities of type inference within
\fexists.

\begin{figure}[h]
  \begin{align*}
    [[double]] &: [[↓(Int → ↑Int)]] \\
    [[map]] &: [[↓(∀α⁺.∀β⁺.↓(α⁺ → ↑β⁺) → List α⁺ → ↑List β⁺)]] \\
    [[len]] &: [[↓(∀α⁺.List α⁺ → ↑Int)]] \\
    [[choose]] &: [[↓(∀α⁺.α⁺ → α⁺ → ↑α⁺)]] \\
    [[id]] &: [[↓(∀α⁺.α⁺ → ↑α⁺)]] \\
    [[auto]] &: [[↓(↓(∀α⁺.α⁺ → ↑α⁺) → (∀α⁺.α⁺ → ↑α⁺))]]
  \end{align*}
  \caption{Polarization of \systemf types}
  \label{fig:polarization-examples}
\end{figure}

\paragraph*{Application of a Polymorphic Function}
  A polymorphic function can be applied to a value of a concrete type. In this
  case, the quantified variables are instantiated to make the application
  type-check. For instance, the polymorphic $[[len]]$ can be applied to a list
  of any positive type, and the resulting type will be inferred by the
  algorithm:
  $$[[· ; Φ ⊢ let x = len([one,two,three]); return x : ↑Int]]$$

\paragraph*{Polymorphic Self-application}
  Unlike predicative systems, \fexists permits self-application of polymorphic
  functions. For instance, $[[id]]$ can be applied to itself resulting in a
  suspended polymorphic computation of type $[[↓(∀α⁺.α⁺ → ↑α⁺)]]$, which, in
  turn, can be applied to itself:
  $$[[· ; Φ ⊢ let x = id(id); let y = x(x); return y : ↑↓(∀α⁺.α⁺ → ↑α⁺)]]$$

\paragraph*{Non-example: Polymorphic Arguments}
The focus of our system is the \emph{local} type inference
\cite{pierce2000:local}, which has certain limitations. In particular, the
instantiation does not happen when the polymorphic term is in the argument
position. For example, one could expect the following to be inferrable:
\mbox{$[[· ; Φ ⊢ let x = map(id, [two, three, nine]); return x : ↑ List Int]]$}.
However, to infer this, $[[id]]$ must be instantiated to \mbox{$[[Int → ↑Int]]$}
which requires the information to be propagated from the neighboring branch of
the syntax tree (\mbox{$[[ [two, three, nine] ]]$}), \ie bypass the locality.

% The annotated version of the same term infers the type successfully, 
% since all the polymorphic variables of the arguments are instantiated:
% $$[[· ; Φ ⊢ let x = map((id : ↓(Int → ↑Int)), [two, three, nine]); return x : ↑ List Int]]$$

\paragraph*{Inferring the Common Supertype}
  The function $[[choose]]$ takes two arguments of the same
  polymorphic type. To apply it to two values of different (concrete) types, 
  the inference algorithm must find the minimal type they can be coerced 
  to---the least common supertype.

@@

\paragraph*{Inferring a general type}
  As mentioned before, the local inference does not instantiate the polymorphic
  types in the argument position.
  Because of that, the application 
  \mbox{$[[let x = choose (id, auto); return x ]]$} cannot infer 
  $[[↑↓(↓(∀α⁺.α⁺ → ↑α⁺) → (∀α⁺.α⁺ → ↑α⁺))]]$.
  However, the inference succeeds and infers a less informative type:
  $$[[· ; Φ ⊢ let x = choose (id, auto); return x : ↑ ∃γ⁻.↓γ⁻]]$$

@@

\paragraph*{Syntactic Conventions}
We always consider terms and types up to alpha-equivalence, and capture-avoiding
substitution. Besides, we assume the quantification is associative,
for example, $[[∀pas.∀pbs.iN]]$ and $[[∀pas,pbs.iN]]$
represent the same type.

\paragraph*{Type Context and Type Well-formedness}

@@

\paragraph*{Missing constructors}
Notice that the language does not have first-class applications: 
their role is played by the applicative let forms, binding 
the result of a \emph{fully applied} function to a variable.
Also notice that the language does not have a type application (i.e. the eliminator of $[[∀]]$) and dually, it does not have \pack (i.e. the constructor of $[[∃]]$).
This is because the instantiation of polymorphic and existential types is inferred by the algorithm. 
In \cref{sec:extensions}, we discuss the way to modify the system to introduce \emph{explicit} type applications.

@@

The inference algorithm for \fexists is \emph{local}: it has a limited scope of
the inference information and does not propagate it between far-apart branches
of the syntax tree. Nevertheless, many difficulties appear in the inference of
the combination of polymorphic and existential types. We now discuss the most
challenging of these difficulties and how they are addressed in the algorithm.

\paragraph*{Subtyping} 
The inference algorithm's complexity mainly lies in the subtyping relation,
which is, in particular, responsible for polymorphic instantiation. It's this
subtyping relation that enables us to apply polymorphic terms in situations
where their specific instantiations are anticipated.

@@

\paragraph*{Equivalence}
The system introduces another complexity: the equivalence induced by subtyping
($[[Γ ⊢ iN ≈ iM]] \defiff [[Γ ⊢ iN ≤ iM]] \text{ and } [[Γ ⊢ iM ≤ iN]]$) is not
straightforward. This relation extends beyond just alpha-equivalent types. As
the neighboring polymorphic quantifiers can be instantiated in arbitrary order.
the equivalence relation must permit the \emph{permutations} of the quantifiers.
For example, we have $[[· ⊢ ∀α⁺.∀β⁺.iN ≈ ∀β⁺.∀α⁺.iN]]$. Similarly, the
equivalence permits the \emph{removal/addition} of the unused quantifiers: $[[·
⊢ ∀α⁺.↑Int ≈ ↑Int]]$.

@@

\paragraph*{Least Upper Bound}
Eventually, the subtyping algorithm reduces the initial subtyping problem to a set of
constraints that need to be resolved. At this moment, another problem arises:
as one variable can occur in multiple constraints, the resolution becomes non-trivial.
In particular, to resolve the conjunction of $[[α̂⁺ ≥ iP]]$ and $[[α̂⁺ ≥ iQ]]$,
we would need to find the least upper bound of $[[iP]]$ and $[[iQ]]$.

@@

To find this least upper bound, we use \emph{anti-unification}---the dual of
unification. While unification finds \emph{the most general instance} of two
given patterns, the anti-unification finds \emph{the most detailed} (under the
restrictions of the system) \emph{pattern} that matches two given types. In the
example above, the anti-unifier of $[[↓↑Int]]$ and $[[↓↑Bool]]$ is $[[↓α̂⁻]]$,
and in \fexists, this unifier is the most specific (in particular because
$[[↓↑α̂⁺]]$ is not considered as a candidate since only \emph{negative} variables are
existentially quantified).

\paragraph*{Impredicativity}
Another difficulty that hinders the constraint resolution arises from the
impredicativity of the system---the ability to quantify over the types that are 
themselves polymorphic. In the impredicative polymorphic system (even
with only $[[∀]]$-quantifiers), the naturally formulated subtyping is
undecidable, which follows from the undecidability of second-order unification.
All the more the undecidability is evident in the presence of existential types
and subtyping constraints. For instance, in \fexists the constraint 
$[[α̂⁺ :≥ ↓↑α̂⁺]]$ would have a surprising impredicative solution $[[α̂⁺]] = [[∃α⁻.↓α⁻]]$.

To maintain decidability in the presence of impredicativity, we carefully
restrict the system using the polarity stratification. One of the main
constraints that we put on subtyping is the invariance of the shift operators:
the subtyping $[[↑iP ≤ ↑iQ]]$ is only allowed if $[[iQ ≥ iP]]$
\emph{and} $[[iP ≥ iQ]]$. These constraints do not trivialize the system: we
still have non-trivial upper bounds, and thus, need to use anti-unification to
find them. However, it prevents such examples as $[[α̂⁺ :≥ ↓↑α̂⁺]]$, and 
as we prove, makes the constraint resolution decidable.
